\documentclass[authoryear]{sigplanconf}

\usepackage[T1]{fontenc}
\usepackage{pslatex}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{verbatim}
\usepackage{pig}
\usepackage{url}
\usepackage{xspace}
\usepackage{listings}
\usepackage{flushend}

\include{macros}

%% Haskell-mode for listings
\lstnewenvironment{code}{\lstset{language=Haskell,
                                 basicstyle=\small\ttfamily,
                                 xleftmargin=1cm}}{}

%% Structure
%\newenvironment{structure}{\footnotesize\verbatim}{\endverbatim}
\newenvironment{structure}{\comment}{\endcomment}

%% Written bits of Structure
\newenvironment{wstructure}{\comment}{\endcomment}

%% Comments
%% \setlength{\marginparwidth}{0.7in}
%% \newcommand{\note}[1]{\-\marginpar[\raggedright\footnotesize #1]%
%%                                   {\raggedright\footnotesize #1}}
\newcommand{\note}[1]{}

%% Syntax
\newcommand{\bind}{\emph{bind}\xspace}
\newcommand{\return}{\emph{return}\xspace}

\begin{document}

\ColourEpigram
%\MonochromeEpigram

\conferenceinfo{ICFP'10,} {September 27--29, 2010, Baltimore, Maryland, USA.}
\CopyrightYear{2010}
\copyrightdata{978-1-60558-794-3/10/09}

%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\title{The Gentle Art of Levitation}


%% Alphabetical ordering.
\authorinfo{James Chapman}
           {Institute of Cybernetics, Tallinn University of Technology}
           {james@cs.ioc.ee}
\authorinfo{Pierre-\'{E}variste Dagand \\ Conor McBride}
           {University of Strathclyde}
           {\{dagand,conor\}@cis.strath.ac.uk}
\authorinfo{Peter Morris}
           {University of Nottingham}
           {pwm@cs.nott.ac.uk}


\maketitle


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Abstract
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\begin{abstract}
  We present a closed dependent type theory whose inductive types
  are given not by a scheme for generative declarations,
  but by encoding in a \emph{universe}. Each inductive datatype arises
  by interpreting its \emph{description}---a
  first-class value in a datatype of descriptions. Moreover, the
  latter itself has a description. Datatype-generic programming thus
  becomes ordinary programming. We show some of the resulting generic
  operations and deploy them in particular, useful ways on the
  datatype of datatype descriptions itself. Simulations in existing
  systems suggest that this
  apparently self-supporting setup is achievable without paradox or
  infinite regress.
\end{abstract}

%\category{CR-number}{subcategory}{third-level}
\category{D.1.1}{Programming Techniques}{Applicative (Functional) Programming}
\category{D.3.3}{Language Constructs and Features}{Data types and structures}

\terms
Design, Languages, Theory



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

Dependent datatypes, such as the ubiquitous vectors (lists indexed by
length) express \emph{relative} notions of data validity. They allow
us to function in a complex world with a higher standard of basic
hygiene than is practical with the context-free datatypes of ML-like
languages. Dependent type systems, as found in
Agda~\cite{norell:agda}, Coq~\cite{coq},
Epigram~\cite{mcbride.mckinna:view-from-the-left}, and contemporary
Haskell~\cite{cheney:gadt, xi:gadt}, are beginning to make themselves useful. As
with rope, the engineering benefits of type indexing sometimes
outweigh the difficulties you can arrange with enough of it.

%Dependent types are an appealing technique for building safer and more
%reliable software. By giving types more expressive power, the
%developer is able to encode more precise invariants in the types. As a
%result, more bugs are caught automatically, during
%type-checking. Because of this benefit, dependently-typed systems have
%flourished, such as Generalized Abstract Data-Types (GADT) in
%Haskell~\cite{spj:gadt}, Agda~\cite{norell:agda},
%Ynot~\cite{morrisett:ynot}, or Epigram~\cite{pigs:epigram}, to name
%but a few.

\begin{wstructure}
<- Describe the problem
    <- Data-types in dependent-type theory
        <- Much more precise
            <- More powerful type-system
            -> Stronger safety guarantees
\end{wstructure}

%In this paper, we will focus on data-types in such systems. Indeed,
%the expressive power of the type-system has a direct impact on
%data-types. Because types can \emph{depend} on terms, our data-types
%can be made more precise. The typical example is vectors, which type
%depend on the size of the vector. Having more precision about
%data-types, we can write safer code: taking the $\CN{head}$ of a
%vector is \emph{ensured} to succeed whenever its type states that it
%is a non-empty vector. This property is automatically enforced by the
%type-checker.

\begin{wstructure}
        <- Equipped with elimination principle
            <- Defining functions over them
            <- Making proofs over them
\end{wstructure}

%Moreover, in total programming systems, such as Agda, Ynot, or
%Epigram, data-types come equipped with an elimination principle: while
%a data-type definition introduces new type formers in the theory, we
%need an eliminator to dispose of them. Looking through the
%Curry-Howard lenses, the elimination principle corresponds to an
%induction principle associated with the data-type. To program over our
%data-types, we rely on their induction principle, guaranteeing the
%well-foundedness of our definition.

\begin{wstructure}
    <- Agda standard library [Nisse file]
        <- x implementations of natural numbers
        <- y implementations of lists
        -> Painful duplication of code and functionality
            <- Types are (slightly) different
                -> Same functions need to be re-implemented 
        -> Crucial need for ``genericity''
\end{wstructure}

The blessing of expressing just the right type for the job can also be
a curse. Where once we might have had a small collection of basic
datatypes and a large library, we now must cope with a cornucopia of
finely confected structures, subtly designed, subtly different. The
basic vector equipment is much like that for lists, but we implement
it separately, often retyping the same code. The Agda standard
library~\cite{nisse:asl}, for example, sports a writhing mass of
list-like structures, including vectors, bounded-length lists,
difference lists, reflexive-transitive closures---the list is
petrifying. Here, we seek equipment to tame this gorgon's head with
\emph{reflection}.

\begin{wstructure}
        /> Coq, Agda: external notion
            <- Not first-class citizen
            -> Cannot "compute" with them
            -> No reflection for data-types (?)
\end{wstructure}

The business of belonging to a datatype is itself a notion
relative to the type's \emph{declaration}. Most typed functional
languages, including those with dependent types, feature a datatype
declaration construct, external to and extending the language for
defining values and programs. However, dependent type systems also
allow us to reflect types as the image of a function from a set of
`codes'---a \emph{universe construction}~\cite{martin-lof:itt}. 
Computing with codes, we expose operations on and
relationships between the types they reflect. Here, we adopt
the universe as our guiding design principle. We abolish the
datatype declaration construct, by reflecting it as a datatype of
datatype descriptions which, moreover, \emph{describes itself}. This
apparently self-supporting construction is a trick, of course, but
we shall show the art of it. We contribute


%As in the simply-typed world, the definition of
%data-types is processed by a meta-theoretical engine, before being
%reifed by extending the type theory with the corresponding type
%formers and elimination principle. Because of this external apparatus,
%data-type definition is not \emph{first-class}: we cannot compute with
%them, such as making new data-types from previous data-types. 
%
%This is a rather harsh limitation, in particular in a
%dependently-typed system. Indeed, reflection~\cite{allen:reflection,
%  gregoire:ring-solver} is at the heart of many dependently-typed
%programming techniques. Not having first-class data-type definitions,
%we have to give up reflection for data-types.

\begin{wstructure}
    <- Dependent types offer new programming techniques
        <- Eg.: universe construction
        /> State of the art haunted by the simply-typed paradigm
            -> Generative
            -> Non reflective
\end{wstructure}

%However, we do not think that we are condemned to such fate. The
%external presentation of data-types is an heritage of the simply-typed
%paradigm. Dependently-typed systems have more to offer. Indeed, new
%programming techniques, unavailable in a simply-typed setting,
%arises. One of them is \emph{universe
%  construction}~\cite{martin-lof:itt}. We shall see how this technique
%help us overcoming the limitations of the standard, non reflective and
%generative presentation of data-types.

\begin{wstructure}
<- State contributions
    <- Closed presentation of data-types 
        -> No generativity requires
        -> Subsuming standard inductive families 
            /> Some popular extensions excluded for now
    <- Descriptions of data-types are first-class 
        <- Self-encoded [Section sec:desc-levitate]
    <- ``generic programming is just programming''
        <- Ability to inspect data-type definition
            -> Write program over them
        <- A generic program works over a class of data-types (???)
            -> Capture this class by common structure
            -> Write a program over this common code
    <- Design a language for generic programming
        -> First serious attempt
            /> except possibly Lisp
                <- ???
\end{wstructure}

%In this paper, we propose a new approach to building data-types in a
%dependent-type theory. Our contributions are the following:

\begin{itemize}
\item a \emph{closed} type theory, extensible only \emph{definitionally},
  nonetheless equipped with a universe of inductive families of datatypes;
\item a \emph{self-encoding} of the universe codes as a datatype in the
  universe---datatype generic programming is just programming;
\item a bidirectional \emph{type propagation} mechanism to conceal
  artefacts of the encoding, restoring
  a convenient presentation of data;
\item examples of generic operations and constructions over our universe,
  notably the \emph{free monad} construction;
\item datatype generic programming delivered \emph{directly},
  not via some isomorphic
  model or `view' of declared types.
\end{itemize}

We study two universes as a means to explore this novel way to equip a
programming language with its datatypes. We warm up with a universe of
\emph{simple} datatypes, just sufficient to describe itself. Once we
have learned this art, we scale up to \emph{indexed} datatypes,
encompassing the inductive families~\cite{dybjer:families,luo:utt}
found in Coq and Epigram, and delivering experiments in generic
programming with applications to the datatype of codes itself.

We aim to deliver proof of concept, showing that a closed theory with
a self-encoding universe of datatypes can be made practicable, but we
are sure there are bigger and better universes waiting for a similar
treatment. Benke, Dybjer and
Jansson~\cite{benke:universe-generic-prog} provide a useful survey of
the possibilities, including extension to inductive-recursive
definition, whose closed-form presentation~\cite{dybjer:axiom-ir,
  dybjer:ir-initial-algebra} is both an inspiration for the present
enterprise, and a direction for future study.

The work of Morris, Altenkirch and
Ghani~\cite{morris:PhD,morris:spf,alti:lics09} on
(indexed) containers has informed our style of encoding and the
equipment we choose to develop, but the details here reflect pragmatic
concerns about intensional properties which demand care in
practice. We have thus been able to implement our work as the basis
for datatypes in the Epigram 2 prototype~\cite{pigs:epigram}. We
have also developed a \emph{stratified} model of our coding scheme
in Agda and Coq\footnote{This model is available at \\
\url{http://personal.cis.strath.ac.uk/~dagand/levitate.tar.gz}}.




%\item We present a basic type-theory and extend it with a universe of
%  finite sets (Section~\ref{sec:type-theory}). We show how coding can
%  be made practical by putting types at work ;
%\item We give a closed presentation of inductive data-types, through a
%  universe of descriptions (Section~\ref{sec:universe-desc}). This
%  first universe has the expressive power of simple inductive
%  types. Being closed, this presentation does not require
%  generativity, hence the type theory remains unchanged when
%  data-types are introduced ;
%\item We present a self-encoding of the universe of description inside
%  itself (Section~\ref{sec:desc-levitate}). As a consequence,
%  description of data-types appears as first-class object in the type
%  theory. We illustrate the benefit of a first-order presentation by
%  implementing a generic catamorphism as well as a generic free monad
%  construction, together with its monadic operations ;
%\item We index the universe of descriptions, to subsume standard
%  inductive families (Section~\ref{sec:indexing-desc}). In this
%  setting, we develop several examples of dependently-typed
%  data-structure and some generic operations over them ;
%\item We have implemented this technology in the Epigram programming
%  language. This is, we believe, the first attempt to design a
%  language for generic programming, Lisp having opened the way. We
%  propose and demonstrate with several examples that generic
%  programming is just programming. Because data-types are described by
%  code, we can finally program with them. As a consequence, generic
%  programs are implemented as functions built from the data-type
%  definition.
%\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The Type Theory
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{paper_type_theory}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% A Universe of simple data-types
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{paper_desc}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Levitating the universe of descriptions
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{paper_desc_levitation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Indexing descriptions
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{paper_idesc}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Discussion
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{paper_discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Conclusion
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \section{Conclusion}
% 
% \begin{wstructure}
% <- System developed in a reasonable theory
%     <- Pi, Sigma, Finite sets
%     /> No assumption about the equality
%     -> Low requirement / high applicability
% \end{wstructure}
% 
% \begin{wstructure}
% <- Formalize a rationnalized presentation of types
%     <- Working directly with codes is not practical
%     -> Bidirectional type-checking
%         <- Type information flows during type checking/type synthesis
%         -> Elaboration turns high-level expressions to low-level terms
%     -> Should not be afraid by codes
% \end{wstructure}
% 
% \begin{wstructure}
% <- Dependently-typed presentation of simple inductive types
%     <- Universe of descriptions
%         <- Based on the specificity of dependent types
%             <- Universe of codes
%             <- Sigma types
%         -> External fix-point and generic induction scheme
%     <- Rationalised by type propagation
%         -> Developer does not see the code
%     <- Self-describing
%         <- Step-by-step exposition
%         -> Minimal extension to the type theory
%             <- Just need fix-point and induction
%         -> Closed presentation of datatype
%             -> Non generative
%         -> datatype is just data
%     <- Generic programming is just programming
%         <- Generic catamorphism
%         <- Generic free monad
% \end{wstructure}
% 
% To introduce our approach, we have presented a universe of
% description. This universe has the expressive power of simple
% inductive types, as found in ML-like languages. Further, we have
% implemented this universe as a self-described object. Hence, for a
% minimal extension of the type-theory, we get a closed, self-describing
% presentation of datatypes, where datatypes are just data.
% 
% \begin{wstructure}
% <- Indexed descriptions for dependent datatypes
%     <- Presented as a slight generalisation of Desc
%         <- Just add indexing
%     <- Develop several examples of datatypes
%         <- Typed syntax
%         <- Constrained datatype a la GADT
%     <- Generic indexed programming
%         <- Indexed free monad
%         <- Substitution
% \end{wstructure}
% 
% To capture dependent datatypes, we generalise our presentation to
% support indexing. The universe of indexed descriptions thus built
% encompasses inductive families. Again, this universe is
% self-described. We have developed several examples of dependent
% datatypes and generic functions over them.
% 
% \begin{wstructure}
% <- All of this without cheating
%     <- Admit a correct stratification
%     <- Terminating
%     <- Strictly-positive types
% \end{wstructure}
% 
% We have presented a self-describing, self-hosted universe for
% datatypes. We have shown the benefit of such approach, by our ability
% to reflect datatypes in the type-theory. This fosters a new way of
% considering generic programming: just as programming. Moreover,
% despite its egg-and-chicken nature, this presentation is free of
% paradox: it has been formalised in Agda, admitting a correct
% stratification.
% 
% \paragraph{Future work:} As such, indexed descriptions do
% not cover several extensions of inductive families. One of them is
% induction-recursion. An interesting question is to locate indexed
% descriptions in the spectrum between inductive families and indexed
% induction-recursion. Another popular extension we plan to consider is
% to allow internal fixpoints and higher-kinded datatypes.
% 
% Also, we have presented a generic notion of syntax with variables,
% thanks to the free monad construction. We would like to explore a
% notion of syntax with binding. Interestingly, introducing internal
% fixpoints or kinds would turn our universe into such syntax with
% binding. Once again, levitation would reveal itself convenient by
% providing generic tools to handle binding.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Appendices
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \appendix
% \section{Appendix Title}

% This is the text of the appendix, if you need one.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Acknowledgments
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\acks

We are grateful to Jos\'{e} Pedro Magalh\~{a}es for his helpful
comments on a draft of this paper. We are also grateful to the Agda
team, without which levitation would have been a much more perilous
exercise. J. Chapman was supported by the Estonian Centre of
Excellence in Computer Science, EXCS, financed by the European
Regional Development Fund. P.-\'{E}. Dagand, C. McBride and P. Morris
are supported by the Engineering and Physical Sciences Research
Council, Grants EP/G034699/1 and EP/G034109/1.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Bibliography
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\bibliography{paper}
\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.
%\begin{thebibliography}{}
%\softraggedright
%\end{thebibliography}

\end{document}
